---
title: "Example MACRO Simulation"
author: "Sean Wu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
header-includes:
   - \usepackage{bbm}
vignette: >
  %\VignetteIndexEntry{MACRO-Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Example MACRO Simulation

## Initialize the Classes Needed for MACRO

In this simulation run we will use the `PfSI` enhanced SIS module of Plasmodium falciparum transmission. Additionally because mosquitoes are not simulated individually in MACRO (they are simulated by integer-valued difference equations), we need the `SimBitePfSI` module to queue infectious bites on human hosts. We also need to initialize the Aquatic Ecology module we wish to use, in this case we choose `Emerge`. To fully initialize all interacting classes we need to run the following code:

```{r}
# attach the MASH package
library(MASH)

# initialize classes for MACRO
MACRO.Humans.Setup()

# setup class methods for PfSI and SimBite modules
# MACRO relies on SimBitePfSI module for adding bites to humans
PfSI.Setup()
SimBitePfSI.Setup()

# MACRO Patch initialization
MACRO.Patch.Emerge.Setup() # 'Emerge' model

```

## Generate MACRO Objects

Now that all classes have had the necessary methods, fields, and pointers added to their `R6Class` Class generator objects, we need to instantiate the specific objects for our simulation run. We can do that as follows on a simple metapopulation tile of 10 patches.

```{r}

# 10 patches in the tile
nPatch = 10

tileParameters = MACRO.Tile.Parameters(N = nPatch,aquaModel = "emerge",aquaPars = list(N=nPatch,lambda=rep(50,nPatch)))
tileParameters$MacroMosquitoPop_PAR$M_density = rep(200,nPatch)
tile = MacroTile$new(MacroTile_PAR = tileParameters)
```

## Simulate MACRO

We are finally ready to run the main simulation. There is only one last step and that is to initialize the human infection for `PfSI` module. We will run for 500 days.

```{r,dev='png'}
tile$init_PfSI(PfPR = rep(0.15,nPatch))
tile$simMacro(5e2)
```

## Analysis of Output

```{r}
pfsiHist = tile$get_HumanPop()$get_History()
plot_PfSI(pfsiHist)

travelHist = tile$get_HumanPop()$get_travelHistory()
# tile$get_HumanPop()$json_travelHistory(con = file(description = "/Users/slwu89/Desktop/OUTPUT/humanTravel.json",open = "wt"))
```
