---
title: "MatrixNotation"
output: pdf_document
---
### Notation 

The following requires developing notation. We let $\vec X$ denote a vector. We will tend to use capital letters for vectors and matrices, and lower case letters for scalars. In the derivations that follow, matrix and vector operations can be a bit confusing. The following guide describes the text notation and its implementation in R:

```{r}
a = .3 
vecX = matrix(c(1,2))
X = matrix(c(1:4),2,2)
Y = matrix(c(5:8),2,2)
```


* If $a$ is a scalar and $\vec X$ is a vector or matrix then let $a \vec X$ denote scalar multiplication. 

* Let $X^t$ denote the transpose of the matrix $X$.

```{r}
t(X)
```

* If $X$ and $Y$ are matrices, then let $X \times Y$ denotes matrix multiplication. 

```{r}
X 
Y
X %*% Y
```
* Let $D(x)$ denote the diagonal matrix, whose on-diagonal elements are $\vec x$.

```{r}
diag(as.vector(vecX))
```

* Let $x \odot \vec y = x \times D(y)$. 

```{r}
X%*%diag(as.vector(vecX))
```



* Let $X$ and $Y$ denote matrices that are the same size. Let $x \cdot y$ denotes elementwise multiplication; if $x_{i,j} \in x$ and $y_{i,j} \in y$ then the corresponding element in $\vec x \cdot \vec y$ is $x_{i,j} y_{i,j}$. 
```{r}
X*X
```

* Let $x$ denote a matrix and $\vec y$ denote a vector, then $x \cdot y$ denotes elementwise multiplication; if $x_{i,j} \in x$ and $y_i \in \vec y$ then the corresponding element in $x \cdot \vec y$ is $x_{i,j} y_i$. 
```{r}
X*as.vector(vecX)
```

Note that $X*Y \neq X \cdot Y$ 

```{r}
X%*%diag(as.vector(vecX))
```

* Let $1/\vec y$ denote the vector whose elements the inverse of the elements of $\vec y$: If $y_i \neq 0 \in \vec y$ then $1/y_i$ is the corresponding element in $1/\vec y$.
```{r}
1/vecX
```

* Let $\vec 1$ be the identity matrix. 

```{r}
diag(1, 2)
```

