---
title: "macro_pfsi_vignette_09_27_2019"
author: "Daniel T. Citron"
date: "9/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MACRO-PfSI VIGNETTE

The purpose of this notebook will be to demonstrate how to use the `macro.pfsi` vignettes, as a way of demonstrating the workflow for setting and running a single simulation.  Eventually we will code up a model which resembles Bioko Island, but for now we just want to make sure the simulator works at all.

```{r Call Libraries}
library(data.table)
library(Rcpp)
library(RcppArmadillo)
library(RcppProgress)
library(devtools)
library(here)
library(withr)
```

## Load macro.pfsi from repo

First, set your path to the local directory of your choosing. You can also use the `reportfactory` library to set up that directory in a structured way.

```{r}
setwd("pick/your/favorite/location")
here()
```


Download the `macro.pfsi` library to a local directory.

```{r Load macro.pfsi from repo}
#This is only for the initial installation
withr::with_libpaths(
  new = here("macro.pfsi"),
  devtools::install_github(repo = "https://github.com/dd-harp/MASH",subdir = "macro.pfsi", force = TRUE)
)

```

## Begin the vignette:

```{r Load Vignette Libraries}

library(macro.pfsi, lib.loc = here("macro.pfsi"))

library(Matrix)
# tidyverse and Hmisc for output visualization; 
# I prefer to use tidyverse but the output can be analyzed with base plots just fine
library(tidyverse)
library(Hmisc)
```

Set the random seed, as setting our initial conditions and running 
```{r Set random seed}
# seed
set.seed(42L)
```

```{r Course of Infection Setup}
# vector of parameters describing course of infection - PfSI
pfsi_pars <- pfsi_parameters(mnChemoprophylaxisPf = 32,
                             PEProtectPf = 1, # % of people the vaccine works in
                             peBlockPf = 1, # % decrease in transmission efficiency of new infections
                             mnPEPf = 200, vrPEPf = 50)

```


```{r Patch set-up}

# set up patches (n is how many patches we have)
n.patch <- 5
# movement matrix for humans is uniform over possible destinations.
move <- matrix(data = 1/(n.patch-1),nrow = n.patch, ncol = n.patch)
diag(move) <- 0
patch_pars <- patches_parameters(move = move,
                                 bWeightZoo = rep(0,n.patch),
                                 bWeightZootox = rep(0,n.patch),
                                 reservoir = rep(F,n.patch),
                                 res_EIR = rep(0,n.patch))
## mosquito parameters
# psi is the movement matrix for mosquitoes, if we let mosquitoes drift from one place to another
psi <- Matrix::sparseMatrix(i = {},j = {},x = 0.0,dims = c(n.patch,n.patch))
diag(psi) <- rep(1,n.patch)

# lambda is the per-day rate that new adult mosquitoes emerge at each patch
# In our model, the abundance of mosquitoes drives the transmission intensity, which in turn drives PR
# (In a more complicated course-of-infection model, the relationship between mosquito abundance, incidence, and PR will be much more complicated)
# The matrix has 1 row for each value of lambda, and 1 column for each day of the year
# This allows us to specify the annual cycles in mosquito emergence rates (but does not yet account for multi-year trends)
# In this example we will have 5 patches, each of which will have a different (increasing) value of lambda
lambda.matrix = t(matrix(c(10,20,30,40,50), nrow = n.patch, ncol = 365))

# Set up mosquito parameters
mosy_pars <- mosquito_rm_conpars(N = n.patch,
                                 lambda = lambda.matrix,
                                 psi = psi,
                                 EIP = rep(11,365),
                                 M = rep(450,n.patch),
                                 Y = rep(0,n.patch),
                                 Z = rep(0,n.patch))
```


```{r Human populations set up}
# humans
patch_sizes <- rep(500,n.patch) # number of people in each patch
pfpr <- rep(0.5,n.patch) # initial malaria prevalence in each patch
n.humans <- sum(patch_sizes) # total number of humans

# sample S or I for each person
# Given that each patch has 50% infected people at the start of the simulation
# we randomly assign each person in each patch to have initial state of infected or susceptible
init_state <- unlist(mapply(FUN = function(n.patch,pr){
  sample(x = c("I","S"),size = n.patch,replace = T,prob = c(pr,1-pr))
},n=patch_sizes,pr=pfpr,SIMPLIFY = F))

# Assign a patch ID to each human (0-indexed, because these variables go into C++)
patch_id <- rep(0:(n.patch-1),times=patch_sizes)

# Assign biting weights to humans
# uniform biting weights; this could follow any density on the positive reals (gamma, log-normal, weibull, etc.)
bweights <- rep(1,n.humans)

# Human parameters - Define parameters to each human
human_pars <- vector("list",n.humans)
for(i in 1:n.humans){
  human_pars[[i]] <- human_pfsi_conpars(id = i-1,
                                        home_patch_id = patch_id[i],
                                        trip_duration = rep(3,n.patch), # assign duration of travel of 3 days to each destination patch to all humans
                                        trip_frequency = 1/20, # assign rate of travel to once every 20 days to all humans
                                        bweight = bweights[i],
                                        age = 20, # relevant if the course-of-infection model had lifetime exposure dependence; not relevant for PfSI
                                        state = init_state[i], # initial state of the ith human
                                        bite_algorithm = 0) # dunno what this one is
}

```


```{r Vaccinations}
# Vaccinations - this is an example of scheduling an intervention

# vaxx_id comes up with a sample of 25 humans to treat
vaxx_id <- sample(x = 0:(n.humans-1),size = n.humans*0.25,replace = F)
# Treat everyone
vaxx_pars <- lapply(X = vaxx_id, FUN = function(id){
  vaccination_pfsi_conpars(id = id, # who to vaccinate
                           t = 500, # vaccinate on day 500
                           treat = T,
                           type = "PE")
})

# Similarly, you can set this list to be empty if you want no interventions
#vaxx_pars <- list()

```



```{r Set up a place to save all of the output files}
log_pars <- list()

# Where to put the human outputs
h_inf <- here("macro.pfsi/vignette_outputs/pfsi.csv")
log_pars[[1]] <- list(outfile = h_inf,key = "pfsi",
                      header = paste0(c("time","patch",unlist(lapply(c("S","I","P"),function(x){paste0(x,c("_visitor","_resident_home","_resident_away"))})),"incidence_resident","incidence_traveller"),collapse = ",")
                    )

# Where to put the mosquito outputs
mosy <- here("macro.pfsi/vignette_outputs/mosy.csv")
log_pars[[2]] <- list(outfile = mosy,key = "mosquito",
                      header = paste0(c("time","state",paste0("patch",1:n.patch)),collapse = ","))
```

```{r Run the simulation}
run_macro(tmax = 1e3,
          human_pars = human_pars,
          mosquito_pars = mosy_pars,
          patch_pars = patch_pars,
          model_pars = pfsi_pars,
          log_streams = log_pars,
          vaxx_events = vaxx_pars,
          verbose = T)
```

# Postprocess and visualize output for analysis
This part you can customize according to your needs - This is the way Sean wrote it up for his example script, although personally I like to use something a little different which doesn't rely on `tidyverse`
```{r Postprocess and Visualize output}
pfsi <- readr::read_csv(h_inf)

pfsi_pr <- pfsi %>%
  select(-ends_with("away")) %>%
  select(-starts_with("incidence")) %>%
  gather(key, value, -time,-patch)

ggplot(pfsi_pr) +
  geom_line(aes(x=time,y=value,color=key)) +
  facet_wrap(. ~ patch) +
  theme_bw()
```

