---
title: "Community detection for coarse-graining clusters of mosquito habitats"
output:
  html_notebook: default
  html_document: default
---

```{r, include = FALSE}
library(igraph)
```


This demo illustrates an artificial landscape of mosquito egg-laying and blood-feeding sites; creates a network of interactions between the different sites; and then uses network community detection to identify strongly interacting clusters of egg-laying sites.

## Generate Artificial Landscape

```{r}
# Create 32 feeding sites, defined by their (x,y) coordinates
xf <- c(rnorm(20, 2, 1), rnorm(10, 8, 1), rnorm(2, 7,.5))
yf <- c(rnorm(20, 3, 1), rnorm(10, 10, 1), rnorm(2,7,.5)) 
# Define the "attractiveness" weight of each feeding site
wf <- rgamma(length(xf), 5, 5)

# Create 29 egg laying sites, defined by their (x,y) coordinates
xl = c(rnorm(5, 2, .5), rnorm(14, 9, .5), rnorm(10, 8, .5)) 
yl = c(rnorm(5, 4, .5), rnorm(14, 4, .5), rnorm(10, 10, .5))
# Define the "attractiveness" weight of each egg laying site
wl = rgamma(length(xl), 5, 5)
```

### Show a plot of the Artificial Landscape
```{r}
par(mar=c(0,6,0,5))
# Plot blood feeding sites in Red
plot(xf, yf, pch = 19, col = "red", ylim = c(-1,12), xlim = c(-1,12), xlab = "", xaxt = "n", ylab = "", yaxt = "n", cex = wf, asp = 1)
# Plot egg laying sites in Blue
points(xl, yl, pch = 8, col = "blue", cex= wl)
```

## Generate Interaction Network
Next we want to express how mosquitoes travel around between these different sites.  This is the interaction strength between the different sites.  Here, we define two kernel functions that quantify how likely it is for a mosquito to travel from a feeding site to an egg-laying site, or from an egg-laying site to a feeding site.  The kernel functions depend on the distances between the two sites and the attractiveness of the target site.  (In practice, though, we can always find other ways to define the interaction strengths between different sites in the habitat.)
```{r}
# Calculate interaction strength between two sites (x,y) and (X,Y)
kerW <- function(x, y, X, Y, w, p=1){
  # Calculate the distance between the two sites (x,y) and (X,Y)
  d = sqrt( (x - X)^2 + (y - Y)^2) 
  # Incorporate the attractiveness of the target site
  exp(-d*p)*w
}

# Vectorize:
# For a single point (x,y), return all interaction strengths between that point and the list of target points (XX, YY)
kerW.i <- function(i, x, y, XX, YY, w, p=1){
  kerW(x[i], y[i], XX, YY, w, p)
}
```

### Apply interaction kernels to the blood-feeding and egg-laying sites
What results is two weighted, non-diagonal, asymmetric matrices, LF and FL.  
* LF represents the interaction strength of traveling from each egg-laying site onto each blood-feeding site.  
* FL represents the interaction strength of traveling from each blood-feeding site onto each egg-laying site.
```{r}
# Interaction strength from each egg-laying site -> each blood-feeding site
LF = sapply(X=c(1:length(xl)), FUN=kerW.i, x=xl, y=yl, XX=xf, YY=yf, w=wf, simplify = "array")
# Interaction strength from each blood-feeding site -> each egg-laying site
FL = sapply(X=c(1:length(xf)), FUN=kerW.i, x=xf, y=yf, XX=xl, YY=yl, w=wl, simplify = "array")

# Transpose, because of R's axis conventions
LF = t(LF)
FL = t(FL)
# If you are confused about the axes of these matrices, try using dim(LF) and dim(FL) to count rows and columns, since there are different numbers of each type of site
```

### Find the interaction strength between the egg-laying sites with themselves
Mosquitoes in our model begin at an egg-laying site, fly to a blood-feeding site, and return to an egg-laying site.  Each egg-laying site, therefore, has the potential to send mosquitoes to each other egg-laying site.

By taking the matrix product between LF and FL, we obtain a matrix that maps egg-laying sites back onto themselves.
```{r}
Q = LF%*%FL
# Row-normalize Q, so that it represents the probability of hopping for one egg-laying site to another egg-laying site
Q = Q / rowSums(Q)
# (To be honest, I'm not sure whether row-normalizing is correct here, but in practice we have total control over how we're defining these matrixes.)
```

### Matrix decomposition into symmetric and non-symmetric components
The matrix Q is weighted and asymmetric, where the asymmetry represents a net flow of mosquitoes from one site to another.  Rather than rewrite Q to merely represent net flow between sites, we decompose Q into a symmetric component C and an asymmetric component N
```{r}
# Asymmetric piece
N = matrix(pmax(0, Q - t(Q)), length(xl), length(xl)) 
# Symmetric piece
C = Q-N
```

### Visualize the symmetric and non-symmetric components of the interaction network, showing how mosquitoes flow from one site to another

```{r, warning=FALSE}
par(mfrow=c(1,2))
par(mar=c(0,0,0,0))
# Symmetric component
plot(xl, yl, cex = .1,
     pch = 16, col = "blue", ylim = c(-1,12), xlim = c(-1,12), 
     xlab = "", xaxt = "n", ylab = "", yaxt = "n")
title("Symmetric component C", line = -1)
for(i in 1:length(xl))
  for(j in i:length(xl))
    segments(xl[i], yl[i], xl[j], yl[j], lwd = C[i,j])
# Asymmetric component, net flow on the landscape
plot(xl, yl, cex = .1,
    pch = 16, col = "blue", ylim = c(-1,12), xlim = c(-1,12), 
    xlab = "", xaxt = "n", ylab = "", yaxt = "n", asp = 1)
title("Asymmetric component N", line = -1)
for(i in 1:length(xl))
  for(j in 1:length(xl)) 
    arrows(xl[i], yl[i], xl[j], yl[j], lwd = N[i,j], length = .05)
```

The main significance of the asymmetric component is that it shows net flows to and from the different egg-laying sites.  We note that the arrows tend to point away from the lower right-hand corner towards each of the other two egg-laying sites. This is because those other egg-laying sites are closer to blood-feeding sites - essentially, mosquitoes leave the lower right-hand corner to move towards the blood-feeding sites and then end up in the other egg-laying sites because they are closer.

## Community structure analysis
We have created an artificial landscape with three artificially-grouped clusters of egg-laying sites.  We have also generated a matrix Q that defines the rate at which mosquitoes jump from one egg-laying site to another, where the rates decrease with distance traveled.

Given that our artificial landscape has three artificially-grouped clusters of sites, and that the interactions between those sites depend on how those clusters are grouped in space, we should be able to use that information to group together the egg-laying sites into clusters that reflect how they are grouped in space.  The matrix Q (as well as each of its components C+N) defines a network of interactions between each of the different sites, and we can use standard network clustering algorithms to define our communities.  Then, we can use the net flow matrix N to find the net flows of mosquitoes between different clusters.

There are two possible strategies for accomplishing this, using Q, C, and N.  

1. We can take the full asymmetric, directed, weighted adjacency matrix Q and run our network clustering algorithms on it - if we pick our clustering algorithms corectly, the results should respect both the strengths of the different connections as well as the net flow through the network.

2. We can run our clustering algorithms on the symmetric component C, and then use the asymmetric component N to determine net flow through the network.  The advantage here is that clustering detection on directed networks is very difficult and most algorithms are optimized to work with symmetric matrices - it may be more straightforward to detect clusters using C and then considering the asymmetric flow separately.

### Convert our matrixes to igraph networks:
```{r}
qnet = graph_from_adjacency_matrix(Q, weighted = TRUE, mode = "directed")
qnet.n = graph_from_adjacency_matrix(N, weighted = TRUE, mode = "directed")
qnet.c = graph_from_adjacency_matrix(C, weighted = TRUE, mode = "undirected")
```

The network community detection literature is enormous, but igraph gives us a few options that are usable with weighted, asymmetric, directed adjacency networks.  

1. Walktrap 
- http://arxiv.org/abs/physics/0512106 
- works with weighted, symmetric, undirected networks (C) 
- this computes community structure using random walks to create a distance metric that is used to distinguish between "nearby" nodes in the same community from "faraway" nodes belonging to a different community

2. InfoMap 
- https://arxiv.org/abs/0707.0609 
- works with weighted, asymmetric, directed networks (N, C, or Q)
- Information-theoretical method for defining community structure. Uses random walks, but this time it searches for a minimum bit sequence required to describe a random walk assuming that it is more likely for random walkers to remain within a community than to leave a community

(There do exist other community detection algorithms that are applicable here, but they do not appear to behave properly with this particular example and so I have not included them.)

### Walk Trap Algorithm
```{r}
walk.c <- cluster_walktrap(qnet.c, weights = E(qnet.c)$weight)
# Plot dendrogram
dendPlot(walk.c, mode = "hclust")
```

The node numbers 1-5 correspond to the lower left-hand cluster, the node numbers 6-19 correspond to the lower right-hand cluster, and the node numbers 20-24 correspond to the upper right-hand cluster of nodes. So using WalkTrap with the C matrix does appear to give us the three communities that we are interested in. 

```{r}
# Visualize the network
plot(qnet.n, 
     vertex.color = walk.c$membership,
     edge.arrow.size = E(qnet.n)$weight*2, edge.width = E(qnet.n)$weight*5
     )
```

The nodes are colored and grouped according to community.  The edges are drawn with directions and weights given by N, so you can get a sense that there is a net flowfrom the orange community (lower right-hand) into the other two communities.

### Coarse graining the network
We can also aggregate together all nodes in each community into a single "metapopulation" node, and sum over all edges between the nodes 

```{r}
# This maps nodes onto their communities
m <- membership(walk.c)
k <- length(unique(m))
coarse.grain <- array(0, dim = c(k,k))
for (i in seq(1,length(E(qnet.n)))) {
  xy <- ends(qnet.n, E(qnet.n)[i]);
  w <- E(qnet.n)$weight[i];
  coarse.grain[m[xy[1]],m[xy[2]]] = coarse.grain[m[xy[1]],m[xy[2]]] + w}
# Define a new network where nodes represent egg-laying sites grouped by community, and edges represent aggregations of edges between the groups of sites
cg.net <- graph_from_adjacency_matrix(coarse.grain, weighted = TRUE, mode = "directed")
# Size aggregate nodes to reflect how many nodes they contain
plot(cg.net, edge.width = (E(cg.net)$weight)^.2, vertex.color = c(1:k), 
     vertex.size = 3*matrix(table(membership(walk.c))))
```

### InfoMap algorithm
Repeat the above using InfoMap!  In this case, the InfoMap algorithm can run using either C or Q (it can run on directed netowrks), so we are going to do both:
```{r}
info.c <- cluster_infomap(qnet.c, e.weights = edge_attr(qnet.c)$weight, modularity = FALSE)
info.q <- cluster_infomap(qnet, e.weights = edge_attr(qnet)$weight, modularity = FALSE)
# Unfortunately InfoMap doesn't create dendrogram-plottable outputs - it's possible that there's a way to adopt the outputs from infomap to a format that supports dendrograms, but I haven't found it yet.
```

```{r}
par(mfrow=c(1,2))
# Visualize the networks
plot(qnet.n, 
     vertex.color = info.c$membership,
     edge.arrow.size = E(qnet.n)$weight*2, edge.width = E(qnet.n)$weight*5
     )
plot(qnet.n, 
     vertex.color = info.q$membership,
     edge.arrow.size = E(qnet.n)$weight*2, edge.width = E(qnet.n)$weight*5
     )
```

So, if we use the full weighted adjacency matrix Q, we end up combining the egg-laying sites in the lower right (a source for mosquitoes) with the egg-laying sites in the upper right (a sink for mosquitoes).  This result could reflect the fact that the majority of mosquitoes from the lower right end up in the upper right, but unfortunately conceals the net flow between those two regions in our landscape.  I'd say that, for this artificial landscape at least, using C to describe the connected clusters and using N to find the net flows between those clusters gives the best clarity in terms of understanding the different relevant components of the landscape.

In practice, it will probably best to use all three of these clustering schemes when we are confronted with landscapes whose clusters are not straightforward and known ahead of time.