---
title: "Modeling Environmental Stochasticity"
output: html_notebook
---

The Poisson distribution is used to measure counts when the expected value, $\mu$ is a constant:
```{r}
mu = 10
N=10^5
hist(rpois(N, mu))
```
For a mixture process, $\mu$ has a distribution. For example, if we let $\mu ~ \Gamma(k, \mu/k)$, as a mixture process, we get the negative binomial distribution: 
```{r}
mu = 10
k = .1
mux = rgamma(N,shape=k, scale=mu/k)
mean(mux)
```
```{r}
hist(rpois(N, mux))
hist(rnbinom(N, mu=mu, size=k), add=TRUE, border="red")
```
To make this as general as possible, we can take any probability distribution function, $E$, to create a mixture process. If we constrain it such that the mean of $E=1$, then we free ourselves up to simulate noise with any form. One way to do this for the negative binomial is to let $E~\mu^{-1} rgamma(shape=k, scale=\mu/k).$ This also allows us to isolate, as a multiplicative factor, the "noise."
```{r}
EE = function(N,mu,k){rgamma(N, shape=k, scale=mu/k)/mu}
env=EE(N,mu,k)
hist(log10(env))
```