---
title: "MBITES / MBDETES formulas for the Length of the Feeding Cycle and Survival through one Feeding Cycle"
author: "David L Smith"
date: "6/8/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Under some restricted conditions on waiting times, MBITES is what we would get if we translated MBDETES into a stochastic model, under the Gillespie algorithm. We can, therefore, use MBDETES to compute expectations for MBITES under these same contitions. 

The utility **mbites2mbdetes** outputs state transitions, with outputs denoted as described in the following. The interval between leaving the post-prandial resting state, $R$ can be computed from the state transition probabilities. Let $P_{X,Y}$ denote the probability of making a transition from state $X$ to state $Y$, and $t_{X}$ the average time spent in $X$. 

### Without Search 

Ignoring search, for the moment, there are two paths out of $R$: refeeding and egg laying, which happen with probability $P_{R,B}$ and $P_{R,O}$. 

The state transitions from $B$ are $B\rightarrow {B,R,D}$. Since we allow $P_{B,B}$, the overall probability of going from $B$ to $R$ (i.e. rather than die, with probability $P_{B,D} = 1-P_{B,B}-P_{B,R}$) is 
$$\hat P_{B,R} = \frac{P_{B,R}}{1-P_{B,B}},$$
with expected waiting time: 
$$T_{B,R} =\frac{t_B}{1-P_{B,B}}.$$

Similarly, the state transitions from $O$ are $B\rightarrow {O,B,D}$, so the overall probability of going from $O$ to $B$ is: 
$$\hat P_{O,B} = \frac{P_{O,B}}{1-P_{O,O}},$$ 
with expected waiting time: $$T_{O,B} = \frac{t_O}{1-P_{O,O}}.$$

So the overall waiting time from $R$ back to $R$ is:
$$\hat P_{R,R} = P_{R,B} \hat P_{B,R} T_{R,B} + P_{R,O} \hat P_{O,B} T_{O,B} \hat P_{B,R} T_{R,B}$$

$$\hat P_{R,R} = P_{R,B}  \left( \frac{P_{B,R}}{1-P_{B,B}} \frac{t_B}{1-P_{B,B}} \right)+ P_{R,O} \left( \frac{t_O}{1-P_{O,O}} + \frac{P_{O,B}}{1-P_{O,O}} \frac{P_{B,R}}{1-P_{B,B}} \frac{t_B}{1-P_{B,B}}\right).$$  

### With Search

With search, we can consider these same paths from $R$ back to $R$ but with the more complicated problem of having repeated attempts to search before successfully completing an attempt. 

The state transitions from $F$ are  $F\rightarrow {F,B,D}$, and from $B$ are $B\rightarrow {F,B,R,D}$. The overall probability of going from $F$ to $B$ is: 
$$\hat P_{F,B}= \frac{P_{F,B}}{1-P_{F,F}}$$
with waiting time: 
$$T_{F,B} = \frac{t_F}{1-P_{F,F}}$$ 
```{r}
hatF2B = function(F2B, F2F){F2B/(1-F2F)}
TF2B = function(tF, F2F){tF/(1-F2F)}
```

The probability of going from $B$ to $R$ is: 
$$\hat P_{B,R} = P_{B,R} + \left(P_{B,B} + P_{B,F} \hat P_{F,B} \right) \hat P_{B,R}$$

$$ \hat P_{B,R} = \frac{P_{B,R}}{1-P_{B,B} - P_{B,F} \hat P_{F,B}} = \frac{P_{B,R}}{1-P_{B,B} - \frac{P_{B,F} P_{F,B}}{1-P_{F,F}}} $$

and: 
$$ T_{B,R} = P_{B,R} t_B + P_{B,B} T_{B,R} + P_{B,F} \hat P_{B,F} \left(T_{B,F} +  T_{B,R} \right) \\ = \frac{P_{B,R}  t_B + P_{B,F} \hat P_{B,F} T_{B,F}}{1-P_{B,B} - P_{B,F} \hat P_{B,F}} 
$$

```{r}
hatB2R = function(F2F,F2B,B2F,B2B,B2R){
  B2R/(1-B2B-B2F*hatF2B(F2F,F2B))
}

TB2R = function(tF,tB,F2F,F2B,B2F,B2B,B2R){
  (B2R*tB + B2F*hatF2B(F2F,F2B)*TF2B(tF,F2F))/
  (1-B2B-B2F*hatF2B(F2F,F2B))
}
```

Similarly, the state transitions from $L$ are  $L \rightarrow {L,O,D}$, and from $O$ are $B\rightarrow {F,L,O,B,D}$.

The overall probability of going from $L$ to $O$ is:
$$\hat P_{L,O}= \frac{P_{L,O}}{1-P_{L,L}}$$
with waiting time: 
$$T_{L,O} = \frac{t_L}{1-P_{L,L}}$$

```{r}
hatL2O = function(L2L, L2O){L2O/(1-L2L)}

TL2O = function(tL, L2L){tL/(1-L2L)}
```

The probability of going from $O$ to $F$ or $B$ is:
$$\hat P_{O,F|B} = P_{O,F} + P_{O,B} + \left(P_{O,O} + P_{O,L} \hat P_{L,O} \right) \left( \hat P_{O,F|B} \right)$$



$$ \hat P_{O,F|B} = \frac{P_{O,F}+ P_{O,B}}{1-P_{O,O} - P_{O,L} P_{L,O}} \\= \frac{ P_{O,F} + P_{O,B}}{1-P_{O,O} -  \frac{P_{O,L} P_{L,O}}{1-P_{L,L}}} $$

and: 
$$ T_{O,F|B} = \left( P_{O,F} + P_{O,B} \right) t_O + P_{O,O} T_{O,F|B} + P_{O,L} \hat P_{L,O} \left(T_{O,L} + T_{O,F|B} \right)
\\= \frac{\left( P_{O,F} + P_{O,B} \right) t_O + P_{O,L} \hat P_{O,L} T_{O,L}}{1-P_{O,O} - P_{O,L} \hat P_{O,L}}
$$

```{r}
hatO2F = function(L2L,L2O,O2F,O2L,O2O){
  O2F/(1-O2O-O2L*hatL2O(L2L,L2O))
}

hatO2B = function(L2L,L2O,O2F,O2L,O2O){
  O2B/(1-O2O-O2L*hatL2O(L2L,L2O))
}

TO2F = function(tF,tO,L2L,L2O,O2F,O2L,O2O){
  (O2F*tO + O2L*hatL2O(L2L,L2O)*TL2O(tL,L2L))/
  (1-O2O-O2L*hatL2O(L2L,L2O))
}

TO2B = function(tF,tO,O2B,O2O,O2F,F2O,F2F){
  (O2B*tO + O2L*hatL2O(L2L,L2O)*TL2O(tL,L2L))/
   (1-O2O-O2L*hatL2O(L2L,L2O))
}
```

### The Length of a Feeding Cycle

We thus have: 
$$T_{R,R} = \sum_{X \in \left\{F,B,L,O\right\}} P_{R,X} \hat P_{X,R} T_{X,R} $$
and survival through a full feeding cycle is: 
$$\hat P_{R,R} = \sum_{X \in \left\{F,B,L,O\right\}} P_{R,X} \hat P_{X,R}$$

```{r}

hatF2R = function(F2B,F2F,B2F,B2B,B2R){
  hatF2B(F2B,F2F)*hatB2R(F2B,F2F,B2F,B2B,B2R)
} 

TF2R = function(tF,tB,F2B,F2F,B2F,B2B,B2R){
  TF2B(tF,F2B,F2F) + TR2B(tF,tB,F2B,F2F,B2F,B2B,B2R)
}

hatO2R = function(F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O){
  hatO2F(L2L,L2O,O2F,O2B,O2L,O2O)*hatF2R(F2B,F2F,B2F,B2B,B2R)+ 
  hatO2B(L2L,L2O,O2F,O2B,O2L,O2O)*hatB2R(F2B,F2F,B2F,B2B,B2R)
}

TO2R = function(tF,tB,tO,tL,F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O){
  O2F*(TO2F(tO,tL,L2L,L2O,O2F,O2B,O2L,O2O) +TF2R(tF,tB,F2B,F2F,B2F,B2B,B2R)) 
  O2B*(TO2B(tO,tL,L2L,L2O,O2F,O2B,O2L,O2O) +TB2R(tF,tB,F2B,F2F,B2F,B2B,B2R))
}

TL2R = function(tF,tB,tO,tL,F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O){
  TL2O(tL,L2L) + TO2R(tF,tB,tO,tL,F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O)
}

hatR2R = function(F2B,F2F,B2F,B2B,B2R,R2F,R2B,R2L,R2O,L2L,L2O,O2F,O2B,O2L,O2O){
  R2F*hatF2R(F2B,F2F,B2F,B2B,B2R)+
  R2B*hatB2R(F2B,F2F,B2F,B2B,B2R)+
  R2F*hatF2R(F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O)+
  R2O*hatO2R(F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O)
}

hatR2R = function(tF,tB,tR,tO,tL,F2B,F2F,B2F,B2B,B2R,R2F,R2B,R2L,R2O,L2L,L2O,O2F,O2B,O2L,O2O){
  tR + 
  R2F*TF2R(tF,tB,F2B,F2F,B2F,B2B,B2R)+
  R2B*TB2R(tF,tB,F2B,F2F,B2F,B2B,B2R)+
  R2F*TF2R(tF,tB,tO,tL,F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O)+
  R2O*TO2R(tF,tB,tO,tL,F2B,F2F,B2F,B2B,B2R,L2L,L2O,O2F,O2B,O2L,O2O)
}
```


