---
title: "Population Rasters of Haiti"
output:
  html_notebook: default
  html_document: default
---

```{r, include = FALSE}
library(sp)
library(rgeos)
library(maptools)
library(MASS)
library(raster)
library(RColorBrewer)
library(GISTools)
library(ggplot2)
library(data.table)
library(lattice)
library(latticeExtra)
library(rasterVis)
```

The first thing we will need is the Haiti data, which can be downloaded from the WorldPop site: http://www.worldpop.org.uk/data/summary/?id=172

I will be using the 2015 population raster, unadjusted, with population per pixel ("ppp"): 172_HTI_ppp_v2b_2015_UNadj/HTI_ppp_v2b_2015_UNadj.tif

## Load in the raster data, using the raster package
```{r}
worldpop <- raster("Haiti_WorldPop/172_HTI_ppp_v2b_2015_UNadj/HTI_ppp_v2b_2015_UNadj.tif")
worldpop.df <- as.data.frame(worldpop, xy = TRUE)
worldpop.dt <- data.table(worldpop.df)
```

```{r}
# Summarize the data set
worldpop
# The dimension of each pixel is 100mx100m, such that Haiti ends up spanning around 250km North-South and 350km East-West, which corresponds to about 3 degrees longitude and 2 degrees latitude
# Each pixel counts the (estimated) number of individuals living there
```

## Load in shape file
```{r, warning=FALSE}
admin2 <- readShapePoly("Haiti_admin2_shapefiles/HTI_A2.shp")
admin2.df <- fortify(admin2)
admin2.dt <- data.table(admin2.df)
```

## Create a raster plot of the population
```{r}
# Define color scheme
mapColors = colorRampPalette(c('#FFEEDC', '#FAA61A', '#5BA7B1', '#2D358E', '#1C1D48'))
mapColors = mapColors(10)

ggplot() +
  # Plot the log of the population data, to give a better sense of the dynamic range of the data
  geom_raster(data=worldpop.dt, aes(x=x, y=y, fill=HTI_ppp_v2b_2015_UNadj)) +
  theme_minimal(base_size=16)  +  # sets grid/background to white and sets base font size
  # Draw district borders
  geom_path(data=admin2.dt, aes(x=long, y=lat, group = group), color="black", size=.05) + 
  # keep proper aspect ratio
  coord_fixed(ratio=1) + 
  # Map title
  ggtitle("Haiti, population distribution") + 
  # Define the range and colors of the legend
  scale_x_continuous("", breaks = NULL) + scale_y_continuous("", breaks = NULL) + #Label axes & remove grid lines
  scale_fill_gradientn(paste("Population"), colours=mapColors, na.value = "white", limits=c(0,1000), seq(0,1000,by=250)) #,

```


## Create a raster plot of the population, with a log scale to show the full dynamic range
```{r}
ggplot() +
  # Plot the log of the population data, to give a better sense of the dynamic range of the data
  geom_raster(data=worldpop.dt, aes(x=x, y=y, fill=log10(HTI_ppp_v2b_2015_UNadj))) +
  theme_minimal(base_size=16)  +  # sets grid/background to white and sets base font size
  # Draw district borders
  geom_path(data=admin2.dt, aes(x=long, y=lat, group = group), color="black", size=.05) + 
  # keep proper aspect ratio
  coord_fixed(ratio=1) + 
  # Map title
  ggtitle("Haiti, population distribution") + 
  # Define the range and colors of the legend
  scale_x_continuous("", breaks = NULL) + scale_y_continuous("", breaks = NULL) + #Label axes & remove grid lines
  scale_fill_gradientn(paste("Log Population"), colours=mapColors, na.value = "white", limits=c(0,3), seq(0,3,by=.5)) #, labels=scale_labels) # defines the color scale / legend
 # can also use "aggregate" function to lower the resolution 

```


### What is the total population of Haiti?
```{r}
# Find the outer border region of Haiti:
border <- gUnaryUnion(admin2)
#  (This operation may be slow)
HTI.pop <- extract(worldpop, border, fun = sum, na.rm = TRUE)
# This is pretty close to the listed CIA worldfactbook population, and also what's currently up on Wikipedia
print(HTI.pop)

```


Sum over the population in each region, and plot
```{r}
a2.names <- admin2$ADM2_NAME
reg.pop.list <- function(x){
  reg <- admin2[admin2$ADM2_NAME == x,]
  reg.sum <- extract(worldpop, reg, fun = sum, na.rm = TRUE)
  return(reg.sum)
}
# This operation may be slow
reg.pops <- lapply(a2.names, reg.pop.list)
```


```{r}
# Create a plot
par(mar=c(0,0,0,0))
shader <- auto.shading(as.numeric(reg.pops), n = 9, cols = brewer.pal(9, "Reds"))
choropleth(admin2, as.numeric(reg.pops), shading = shader)
choro.legend(-74.6, 19.9, shader)
```

